Applying Pose Estimation to Predict the Outcome of Basketball Shots

### 2022-11-28 Papers/Report/report.tex
Add final report files and revise introduction. Add abstract and work on related works

%%%%%%%%% TITLE - PLEASE UPDATE
\title{The Shot Predictor}

@@

%%%%%%%%% ABSTRACT

%%%%%%%%% BODY TEXT
\section{Problem Description}
\label{sec:intro}

In basketball, the ability of a player to effectively shoot the basketball typically comes down to the player’s shooting form. While the form of the best shooters tend to look different, they all typically use the same fundamentals. In our project, we will attempt to capture these fundamental aspects of a player’s shooting form and attempt to predict the outcome of a shot using these features. Research has been done on extracting features from a player’s movement to classify the action a player is performing (shooting, dribbling, etc.) \cite{Basketball}, but we would like to focus our energy on feature extraction from the shooting motion using pose estimation \cite{OpenPose}, object detection \cite{YOLO}, and possibly other methods to extract feature descriptors of a shot and attempt to identify it as a make or a miss.\\
\indent This problem is a particularly nontrivial application of pose estimation for two main reasons. The first being that there are multiple stages to
a basketball shot that need to be taken into account. From dipping the ball to waist level, to the motion of bringing the ball to eye level, to releasing the ball, each
plays a significant role in the outcome of a shot, so each stage needs to be taken into account. The second reason is the variability of the average shot. It can be argued that no two shots are identitical, when taking into account the speed of the shot
and different shooting forms. This will require us to develop a method of capturing all the motions of a shot in a way that is invariant to the shot's speed and overall form.

%-------------------------------------------------------------------------
%------------------------------------------------------------------------
\section{Related Works}
\label{sec:formatting}

@@

Pose Estimation is a critical topic in computer vision that will intersect with our goal of trying to accurately capture the motion and actions of a person taking a shot in basketball. Pose estimation, in the context of 2D videos of humans, is the problem of localizing anatomical keypoints or joints in a frame by frame video or image \cite{OpenPose}. To fulfill our goal of predicting the outcome of a basketball shot, it will be critical to assess the form of a player who's taking a shot - where form can be decomposed into various classifications of joints in space. Pose estimation methods can be categorized into bottom-up or top-down methodologies. Bottom-up methodologies start by estimating keypoints and body joints first, and then these points are clustered to form poses. In contrast, 
top-down methodologies of pose estimation first run a person detector before decomposing each person into their respective body joints within detected bounding boxes \cite{Viso2}. Computational complexity is a major consideration for landmark pose estimation algorithms, and modern SOTA pose estimation algorithms deploy deep learning and CNNs to improve compuational overhead and speed \cite{OpenPose}. We list some examples of prevalent and SOTA pose estimation models that have been employed and researched below.
%
\begin{quotation}
  OpenPose: The first multi-person realtime 2D pose estimation system that uses a bottom-up approach that implements nonparametric representation to associate human keypoints and body parts with an individual in an image \cite{OpenPose}. \\
  \newline
  \indent DeepPose: SOTA pose estimation method that uses DNNs to classify human body joints through the usage of cascading DNN regressors that produce high precision pose estimates \cite{DeepPos}. \\
  \newline
  \indent AlphaPose: Multi-person SOTA realtime pose estimation system that outperforms OpenPose in AP score and has a high mAP score \cite{Alpha}. \\
  \newline
  \indent DeepCut: Proposes an approach to solving issues in both pose estimation and detection by using a partitioning and labeling formulation of a set of CNN part detectors \cite{DeepCut}.
\end{quotation}
%

@@

Collecting and analyzing basketball player's posture data is important and 
highly sought after in the scientific basketball community in order to help
coaches improve their training plan, and help players maximize their training 
outputs. As it stands, the current technology faces some combination of 
high inefficiency or high error rates. Rather than using wearable, inertial sensor
recognition equipment, which tends to be expensive and not applicable to how a 
real game is played, it was decided to use posture recognition via video instead \cite{Basketball}.
In order to improve the athlete's training effect, a gesture recognition method 
based on feature extraction and machine learning was implemented in order to 
acquire the upper limb and lower limb features needed to classify a posture.  
The action posture data of players was collected through image feature extraction
methods, and with the method of feature selection, accurate classification and recognition
of basketball shooting gestures can be made \cite{Basketball}. This paper is similar to our project because 
it attempts to take and encode data about motion of a player through pose estimation to 
make predictions, but our project solely focuses on the motion of a shot, and predicting 
the outcome of the shot.

>>>>

%%%%%%%%% TITLE - PLEASE UPDATE
\title{Applying Pose Estimation to Predict the Outcome of Basketball Shots}

@@

%%%%%%%%% ABSTRACT
\begin{abstract}
  In this report, we present an end-to-end pipeline that includes a pose estimator as well as 
  a classification model that can effectively predict the outcome of a free throw. Our training data was
  self-generated by recording hundreds of clips of our test subject shooting free throws of various forms
  and recording the outcome. We explore different ways to generate feature vectors for inputs to our model as 
  well as multiple classification models to produce the best performing pipeline.
\end{abstract}

%%%%%%%%% BODY TEXT
\section{Introduction}
\label{sec:intro}

In basketball, the ability of a player to effectively shoot the basketball typically comes down to the player’s shooting form. While the form of the best shooters tend to look different, they all typically use the same fundamentals. In our project, we will attempt to capture these fundamental aspects of a player’s shooting form and attempt to predict the outcome of a shot using these features. Research has been done on extracting features from a player’s movement to classify the action a player is performing (shooting, dribbling, etc.) \cite{Basketball}, but we would like to focus our energy on feature extraction from the shooting motion using pose estimation \cite{OpenPose}, object detection \cite{YOLO}, and possibly other methods to extract feature descriptors of a shot and attempt to identify it as a make or a miss.\\
\indent This problem is a particularly nontrivial application of pose estimation for two main reasons. The first being that there are multiple stages to
a basketball shot that need to be taken into account. From dipping the ball to waist level, to the motion of bringing the ball to eye level, to releasing the ball, each
plays a significant role in the outcome of a shot, so each stage needs to be taken into account. The second reason is the variability of the average shot. It can be argued that no two shots will ever be identical due to the imperfect nature of humans.
Therefore, it is necessary for our feature representation of a shot to be invariant towards miniscule changes
in shot form and focus more on fundamental differences.
%-------------------------------------------------------------------------
%------------------------------------------------------------------------
\section{Related Works}
\label{sec:formatting}

@@

Pose Estimation is a critical topic in computer vision that will intersect with our goal of trying to accurately capture the motion and actions of a person taking a shot in basketball. Pose estimation, in the context of 2D videos of humans, is the problem of localizing anatomical keypoints or joints in a frame by frame video or image \cite{OpenPose}. To fulfill our goal of predicting the outcome of a basketball shot, it will be critical to assess the form of a player who's taking a shot - where form can be decomposed into various classifications of joints in space. Pose estimation methods can be categorized into bottom-up or top-down methodologies. Bottom-up methodologies start by estimating keypoints and body joints first, and then these points are clustered to form poses. In contrast, 
top-down methodologies of pose estimation first run a person detector before decomposing each person into their respective body joints within detected bounding boxes \cite{Viso2}. Computational complexity is a major consideration for landmark pose estimation algorithms, and modern SOTA pose estimation algorithms deploy deep learning and CNNs to improve compuational overhead and speed \cite{OpenPose}. We list some examples of prevalent and SOTA pose estimation models that have been employed and researched below.
%
\begin{quotation}
  OpenPose: The first multi-person realtime 2D pose estimation system that uses a bottom-up approach that implements nonparametric representation to associate human keypoints and body parts with an individual in an image \cite{OpenPose}. \\
  \newline
  \indent DeepPose: SOTA pose estimation method that uses DNNs to classify human body joints through the usage of cascading DNN regressors that produce high precision pose estimates \cite{DeepPos}. \\
  \newline
  \indent AlphaPose: Multi-person SOTA realtime pose estimation system that outperforms OpenPose in AP score and has a high mAP score \cite{Alpha}. \\
  \newline
  \indent DeepCut: Proposes an approach to solving issues in both pose estimation and detection by using a partitioning and labeling formulation of a set of CNN part detectors \cite{DeepCut}.
\end{quotation}

Pose estimation attempts to detect the location of 17 keypoints on a human body, as seen in \ref{fig:pose}. These keypoints include key joints such as
the knee, elbow, shoulder, and wrist, which can be vital in determining the form of a free throw.
\begin{figure}[h]
  \centering
  \includegraphics[width=0.25\textwidth]{imgs/openpose.png}
  \caption{Pose estimation keypoints}
  \label{fig:pose}
\end{figure}
%

@@

The application of pose estimation in basketball is not a new concept. Collecting and analyzing basketball player's
posture data is an important facet of the scientific basketball community in order to help maximize training outputs.
For example, pose estimation was used in combination with classification models to predict the action a player is performing
in a video \cite{Basketball}. This is fairly similar to our project because it attempts to create feature vectors in video frames
using pose estimation to encode data about the motion of a player and use these features to make a prediction. However, our project
solely focuses on the motion of a player's shooting form and predicting the outcome of the shot.
